---
title: Using the Research Data Warehouse
teaching: 15
exercises: 15
---

```{r, echo=FALSE}
# Source the external configuration script
source("load_config.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Understand how to use Newcastle University's Research Data Warehouse (aka, RDW and Campus Filestore) with Comet HPC

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How do I transfer files to (and from) the cluster?
- What is the best way to back up research data?

::::::::::::::::::::::::::::::::::::::::::::::::::


## Transferring files to and from Campus Storage for Research Data (RDW)
RDW (Research Data Warehouse) is mounted on Comet at `/rdw` so you can access it just like any local filesystem.
Research project owners can request their own share on RDW for safe storage of research data.
Although RDW a separate physical system, it's located in the same data centre as Comet and connected via fast ethernet.
You can use `cp` and `rsync` to transfer data to RDW in the same way as copying to any other directory on Comet.  
RDW is intended for data storage and NOT suitable for interactive use or software installation.  
Working data should be in your home or project directory.
User installed software should be in your home directory.
We can practice making a backup of the amdahl software we uploaded in the last episode.

### Using cp to copy to RDW
Because `/rdw` is a mounted filesystem, we can use `cp` instead of `scp`.  
Let's make our own directory inside the RDW share belonging to `r config$remote$project`:

```bash
`r config$remote$prompt` pwd
```
```output
/mnt/nfs/home/`r config$remote$user`
```
```{r, echo=FALSE, results='asis'}
snippets('rdw/create-dir.Rmd')
```

```output
/rdw/04/rse-training/`r config$remote$user`
```
```bash
[user@cometlogin02(comet) `r config$remote$user`]$ ls
```
```output
example-job.sh
```

### Using rsync to copy to RDW

As you gain experience with transferring files, you may find the `cp` and `scp`
commands limiting. The [rsync](https://rsync.samba.org/) utility provides
advanced features for file transfer and is typically faster compared to both
`scp` and `sftp` (see below). It is especially useful for transferring large
and/or many files and creating synced backup folders.
The syntax is similar to `cp` and `scp`.  Rsync can be used on a locally mounted filesystem or a remote filesystem.

Transfer *to* RDW from your home directory on Comet

#### Try out a dry run:

```{r, echo=FALSE, results='asis'}
snippets('rdw/rsync-amdahl-dry.Rmd')
```

#### Run ‘for real’:

```{r, echo=FALSE, results='asis'}
snippets('rdw/rsync-amdahl.Rmd')
```
and check the result

```{r, echo=FALSE, results='asis'}
snippets('rdw/rsync-amdahl-result.Rmd')
```

    
:::::::::::::::::::::::::::::::::::::::::  callout

## Common options for `rsync`
The usual format for an `rsync` command is:

`rsync -av source/directory/path destination/directory/path`

The `-a` (archive) option is equivalent to -rlptgoD. It is a quick way of saying you want to recurse through directories and to preserve almost everything, including permissions. 
Use `man rsync` or `rsync --help` to find out more. 
Because permission groups on RDW are set outside of Comet, we use a subset of `-a`

For Comet and RDW, replace `-av` with `-rltv`  
`-r` = recurse through subdirectories    
`-l` = copy symlinks    
`-t` = preserve timestamps   
`-v` = verbose

::::::::::::::::::::::::::::::::::::::::::::::::::


## Large data copies
When copying large amounts of data, rsync really comes into its own. When you're copying a lot of data, it's important to keep track in case the copy is interrupted.  Rsync is great because it can pick up where it left off, rather than starting the copy all over again.  It's also useful to output to a log so you can see what was transferred and find any errors that need to be addressed.

### Fast Connections
Transfers from Comet to RDW don’t leave our fast data centre network.  If you're using rsync with a fast network or disk to disk in the same machine:

- DON'T use compression `-z`
- DO use `--inplace`

Why?  compression uses lots of CPU, Rsync usually creates a temp file on disk before copying.  For fast transfers, this places too much load on the CPU and hard drive.    
`--inplace` tells rsync not to create the temp file but send the data straight away.  It doesn’t matter if the connection is interrupted, because rsync keeps track and tries again.  Always re-run transfer command to ensure nothing was missed.  The second run should be very fast, just listing all the files and not copying anything.

### Slow Connections
For a slow connection like the internet:
 
- DO use compression `-z` 
- DON’T use `--inplace`.  


:::challenge
##  Large Transfer to RDW
RDW has a super-fast connection to Comet, which means that it takes more resource to compress and un-compress the data than it does to do the transfer.
What command would best for backing up a large amount of data from Comet to RDW?

:::solution
```bash
[`r config$remote$user`@login01 ~]$ rsync -rltv --inplace --size-only --stats DataDir /rdw/04/rse-training/`r config$remote$user`/
```

--inplace - saves resources by not creating temporary files
--size-only - saves time by only checking whether a file's size has changed (and not its last-modified time) 

 
:::

see `man rsync` and https://rsync.samba.org/ for more examples
--delete is an option that is very useful for tidying up when files have been duplicated.  However it should be used with care!  
Perhaps a collaborator has placed additional files in the directory you are syncing to.  
Use --dry-run --progress --stats to check before you run
Accidental deletions on RDW can be rolled back using Windows File Explorer.  Log a ticket with NUIT for help with rollback.

:::

:::challenge
## add a dry run and a log file
 
:::solution
Try out a dry run:
```bash
rsync --dry-run -rltv --inplace --itemize-changes --progress --stats  --size-only /nobackup/myusername/source /rdw/path/to/my/share/destination/ 2>&1 | tee /home/myusername/meaningful-log-name.log1
```
Run ‘for real’:
```bash
rsync -rltv --inplace --itemize-changes --progress --stats  --size-only /nobackup/myusername/source /rdw/path/to/my/share/destination/ 2>&1 | tee /home/myusername/meaningful-log-name.log2
```

- `--inplace  --size-only` speed up transfer and prevent rsync filling up space with a large temporary directory    
- `--itemize-changes --progress --stats` for more informative output    
- Remember `|` from the Unix Shell workshop?    
 `| tee` sends output both to the screen and to a log file    
- All the arguments can be single letters like `-v` or full words like `--verbose`. Use `man rsync` to craft your favourite arguments list.

:::
:::


:::::::::::::::::::::::::::::::::::::::: keypoints

- `cp` and `rsync` transfer files between RDW and HPC.
- try a dry-run of rsync to avoid accidental duplications or deletions
- re-run large rsync commands to confirm success
- output to a log to keep a record
- group permissions on RDW can't be changed from linux
- RDW shares have a pre-set 'modify' group of campus users
- some RDW shares have a pre-set 'read' group of campus users
- RDW has a roll-back feature in case of accidents

Find out more about where to store data on Comet:
https://hpc.researchcomputing.ncl.ac.uk/dokuwiki/doku.php?id=started:filesystems


::::::::::::::::::::::::::::::::::::::::::::::::::
