---
title: Parallelising with Job Arrays
teaching: 15
exercises: 5
---

::::::::::::::::::::::::::::::::::::::: objectives

- Prepare a job submission script for an array job.
- Launch a job to be executed in parallel over several nodes

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What are job arrays?
- What benefits do job arrays bring?
- What type of jobs would benefit from job arrays?

::::::::::::::::::::::::::::::::::::::::::::::::::


::::instructor
Download the archive of data and scripts using wget https://raw.githubusercontent.com/NewcastleRSE-Training/hpc-intro-comet/refs/heads/main/episodes/files/array-jobs.tar.gz
::::

## Job Arrays for High Throughput

Parallel computing is a technique used to divide big tasks into smaller ones that can be solved simultaneously. 
Parallelism can be accomplished in different ways; the best approach depends on the tasks that need doing as well as the algorithms implemented to perform these tasks.

One way of implementing parallel computing is to distribute a job across multiple processors. 
This is usually accomplished by using the Message Passage Interface (MPI) which is a standardised way for CPU cores to communicate with one another while working together on a task. 
Software has to be written specifically to utilize MPI to take advantage of this.

Another form of parallel computing is an array job. This type of job is advantageous where the same software has to be run across several files. 
An example of this would be in bioinformatics where the same workflow has to applied to a set of files that contain data for different samples. 
There is no need for the different jobs to "talk" to one another while they run. The advantage lies in the fact that the jobs can run in parallel. 
One could potentially run such processes manually across different computers but imagine having a hundred files and each taking two hours to complete. 
You can run them in series which would take two hundred hours or you can manually start them across, say, four computers which would mean it would take 25 hours. 
But it would take you some time to start all these jobs if you do it manually. To complicate matters things will quite often go wrong and workflow won't complete in which case you have to first notice this, correct the problem and then restart it all.

Array jobs are controlled by the Slurm scheduler. You will need only one set of scripts to which you supply a list of files. 
Slurm will automatically distribute the jobs across available nodes. If any of the jobs fail you can easily restart only those failed jobs.

## A case of counting words

Peter, a linguistics researcher, wants to investigate changes in language over time by comparing how often words are used in various texts. 
The data consists of several books from the Gutenberg project as text files:

|Filename|Book name|
|---|---|
|data.1|The collected works of Shakespeare|
|data.2|Geoffrey Chaucers Cantebury Tales |
|data.3|Moby Dick by Herman Melville|
|data.4|Homers Odyssey|

Peter has been doing this work on their laptop using a programme called `word-freq.sh` but it's taking far too long so they have decided to move their work to HPC in order to get through the processing more quickly.

### Preparing a directory to work in

First, create a directory in a shared area so that your collaborators can access your work:


```{r, echo=FALSE, results='asis'}
snippets('array/set-up-work-dir.Rmd')
```

Gather the scripts and data into a working directory:

```{r, echo=FALSE, results='asis'}
snippets('array/download-extract-data.Rmd')
```

### Checking the script runs as expected

Create a small data file to test our script:

```bash
`r config$remote$prompt` nano test-data.txt
`r config$remote$prompt` cat test-data.txt
```

```bash
This is a small file - it will be very useful for trying out our script.
Some words are repeated in this file
- we can look for repeated words
and count them (to see which words are repeated most often).
```

To test the script we will run it on the login node. Remember, never do this with resource intensive script. You could even run the script on your laptop or desktop if it uses Linux or Mac. This specific script will not work on Windows as not all the commands in the script are available on the Windows operating system.

```bash
bash word-freq.sh test-data.txt
```

To the results, type the output of the script to screen:

```bash
cat analyzed_test-data.txt
```

You should get something like this:

```
      1 a
      1 and
      1 be
      1 can
      1 count
      1 in
      1 is
      1 it
      1 look
      1 most
      1 often
      1 our
      1 out
      1 script
      1 see
      1 small
      1 some
      1 them
      1 to
      1 trying
      1 useful
      1 very
      1 we
      1 which
      1 will
      2 are
      2 file
      2 for
      2 this
      3 repeated
      3 words

```

### Create a submission script

Once we have proved that the script runs without a problem we can write a script that can be submitted to Slurm. 
We will check our submission script first by using our test data, rather than trying to run with a large dataset.
Using `nano`, create a script called `job_single_word-freq.sh` containing the following:

```bash
#!/bin/bash

#SBATCH --partition=short_free
#SBATCH --account=comet_training
#SBATCH --job-name=word-freq_single-test1
#SBATCH --nodes=1
#SBATCH --tasks=1
#SBATCH --cpus-per-task=1

echo "Starting word frequency script"
bash word-freq.sh test-data.txt
echo "Finished word frequency script"
```

::::instructor
The complete script can be downloaded from: https://raw.githubusercontent.com/NewcastleRSE-Training/hpc-intro-comet/refs/heads/main/episodes/files/job_single_word-freq.sh
::::




# Andrew's version
Parallel computing allows multiple computational tasks to execute
simultaneously in order to reduce execution time for a task, or
increase throughput for multiple tasks. Depending on the application,
the workload may be divided into cooperating subtasks that communicate with
one another, or into independent tasks that execute separately.

One common approach to parallel computing is to distribute computation across
multiple processes that cooperate by exchanging information during
execution using the Message Passing Interface. This is the "MPI" that we
saw as a dependency of the `amdahl` application.

Not all workloads require processes to cooperate. Many scientific workflows
are made up of independent jobs. For these cases, `r config$sched$name`
provides Job Arrays, allowing many similar jobs to be submitted and managed together.

For instance,
you might need to run the same task on several independent input
files, or you may have multiple serial tasks that take some parameter,
and you need to explore several values of the parameter.
Workflows made up of these independent elements are 
also sometimes called "high-throughput" computing.

Operationally speaking, a Job Array is a collection of related batch
jobs submitted using a single job script.

:::::::::::::::::::::::::::::::::::::::  challenge

What distinguishes workloads that are suitable for job arrays from those
that require traditional parallel programming?
Describe some examples.

:::::::::::::::  solution

Tasks appropriate for array jobs are "high-throughput", where
the same thing needs to be done many times, possibly over a set
of parameters, but where each task is independent of the others.

For example, running the same statistical analysis on a large
number of independent input files is a good candidate for an
array solution.

Parallel tasks which have interactions between the various
parallel processes need to communicate between processes at
run-time, and are not appropriate for job arrays.

For example, most parallel scientific codes that run in
parallel have a requirement to communicate between parallel
elements at run-time, and are not appropriate for job arrays.

Similarly, serial tasks which only need to be run once do not
benefit from parallelism. Aggregating unrelated tasks into an
array merely for the sake of grouping does not make sense.

:::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::::::

We have already seen how to submit a single serial job to the
Slurm scheduler.

Array jobs differ in character, but are specified in a similar manner.
The batch file describes the array to the system, and at submission
time, all of the array elements are submitted together. They may
dispatch at different times, and may or may not run concurrently.

Within
the execution environment of an array element, a number of environment
variables will be set, which can be used at run-time by the
array elements to identify the size of the array, and their
position within it, allowing them to select the correct input
file. These jobs will be dispatched as their resource requirements
and priorities are met, and they may or may not run concurrently.

By way of review, we will construct a job submission script that
runs the `amdahl` executable with the default arguments.

Using `nano`, create a script called
`job_single_amdahl.sh` containing the following:

```{r, echo=FALSE, results='asis'}
snippets('array/job_single_amdahl.Rmd')
```


:::::::::::::::::::::::::::::::::::::::  challenge

How would you submit the script to Slurm for execution?

:::::::::::::::  solution

  
```bash
sbatch job_single_amdahl.sh
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Array Job Syntax


To specify an array job, you only need to add a single
directive to your batch file, and then adapt your run command
to take advantage of the information provided by the environment
variables.

The relevant array directive has this format:

```bash
`r config$sched$comment` `r config$sched$flag$array`=<array-spec>
```

The `<array-spec>` above is a place-holder for specifying the
size and extent of the array. The specification will resolve to
set of integers, which will index the job array elements.

For a simple example, an array specification of `1-4` means the
system should create four array elements, numbered consecutively
from one through four.

You can also specify a comma-separated set of numbers, such
as `1,3,5`, or you can specify a stride, for example by specifying
`1-10:2` (which is equivalent to `1,3,5,7,9`).

In addition to these, you can also specify a limit on the number
of array elements that will run concurrently, using the `%` sign.
An example of this, building on what we saw before, would be
to specify `1-10:2%4`, which will create five array elements
with indices 1, 3, 5, 7, and 9, and run at most four of them
at a time until they are all complete.

When an array element job is running, the run-time environment
will include some special environment variables, the most important
of which is `SLURM_ARRAY_TASK_ID`, which specifies the index
of the current instance.  There are other environment variables
which tell you the full size of the array, and the starting and
ending indices. As we have seen, because there is a fairly rich
syntax for specifying arrays, it is not straightforward to
infer the size of the array from the high and low indices.

There are also some file-name patterns you can use to control
where your executable reads and writes data. The most important
of these is the `%a` pattern, which corresponds to the index
of the current array element, similarly to `SLURM_ARRAY_TASK_ID`.

::::: challenge


Write a batch script to run the `amdahl` program as an array job with
4 independent jobs, to examine the effects of jitter.

Use the directives discussed above to specify the array elements,
and to direct the output of each array element to a separate file.

:::: solution

```{r, echo=FALSE, results='asis'}
snippets('array/job_array_amdahl.Rmd')
```

::::

:::::



::::: challenge

Write a batch script to call the word-freq.sh as an array job with 4 parallel jobs to process all 4 text files (job_array_word-freq.sh). To do this you will need the directive `#SBATCH --array=1-4`. When using this directive, each job will be given a job number. In this case it will be job numbers one to four. While running the script for a specific job number, that number will be available in an environment variable called `${SLURM_ARRAY_TASK_ID}`.

:::: solution

```bash

`r config$remote$bash_shebang`

`r config$sched$comment` `r config$sched$flag$queue`short_free
#SBATCH --account=comet_training
`r config$sched$comment` `r config$sched$flag$name`makefreq
`r config$sched$comment` `r config$sched$flag$nodes`1
#SBATCH --array=1-4
`r config$sched$comment` `r config$sched$flag$cpu`1

# Do a word frequency analysis of each of the following
# data sets simultaneously:
#
# data.1 - The collected works of Shakespeare
# data.2 - Geoffrey Chaucers Cantebury Tales 
# data.3 - Moby Dick by Herman Melville
# data.4 - Homers Odyssey
#
# We should be able to process all four data sets in the same
# time it took to process just the first.

echo "Starting word frequency script"
bash word-freq.sh data.${SLURM_ARRAY_TASK_ID}
echo "Finished word frequency script"
```

::::

:::::


:::: instructor

You can download the script from https://raw.githubusercontent.com/NewcastleRSE-Training/hpc-intro-comet/refs/heads/main/episodes/files/job_array_word-freq.sh

::::




:::::::::::::::::::::::::::::::::::::::: keypoints

- Parallel programming allows applications to take advantage of parallel hardware.
- The queuing system facilitates executing parallel tasks.
- Parallel computing allows applications to distribute the workload over several CPUs or nodes
- There are multiple parallelization strategies that are generally supported by resource managers.
- Array parallel jobs are suitable for independent runs of the same executable with varying inputs or outputs. 


::::::::::::::::::::::::::::::::::::::::::::::::::
